\Rpackage{MSnbase}: labelled MS2 data pre-processing, visualisation and quantification.

### 2011-04-26 inst/doc/MSnbase-demo/MSnbase-demo.Rnw
Quantification by trapezoidation now has strict argument. Removed many warnings in quantify. Updated vignette with illustrative plots for quantitation methods and spectrum pre-processing. See NEWS for more details.
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MSnbase@55087 bc3139a8-67e5-0310-9ffc-ced21a209358

Unlike the name might suggest, the \Rfunction{removePeaks} method does 
not actually remove peaks from the spectrum; they are set to 0. This 
can be checked using the \Rfunction{peaksCount} method, that returns 
the number of peaks (including 0 intensity peaks) in a spectrum.
To effectively remove 0 intensity 
peaks from spectra, and reduce the size of the data set, one can use 
the \Rfunction{clean} method.

@@

Another useful manipulation method is \Rfunction{trimMz}, that takes as 
parameters and \Robject{MSnExp} (or a \Robject{Spectrum}) and a numeric 
\Robject{mzlim}. MZ values smaller then \texttt{min(mzlim)} 
or greater then \texttt{max(mzmax)} are discarded. This method is 
particularely useful when one wants to concentrate on a specific MZ range, 
as for reporter ions quantification, and generally results in substantial 
reduction of data size. Compare the size of the full trimmed experiment to 
the original \Sexpr{sz} Mb.
\label{timMz-example}

@@

Quantitation is performed on specific peaks in the spectra, that are specified 
with an \Robject{ReporterIon} object. A specific peak is defined by it's expected 
\Robject{mz} value and is expected to be found within 
\Robject{mz} $\pm$ \Robject{width}. If the peak reaches outside, a warning will 
be issued. If no data is found, \Robject{NA} is returned.

@@

The \Rfunction{quantify} method takes an experiment, 
a character describing the quantification 
\Robject{method} and the \Robject{reporters} to be quantified as parameters. 
Additionally, a progress bar can be dispaying when setting 
the \Robject{verbose} paramter to \Robject{TRUE}.
Three quantification methods are implemented: 
\Robject{trapezoidation} returns the area under the peak of interest, 
\Robject{max} returns the apex of the peak and 
\Robject{sum} returns the sum of all intensities of the peak.

The \Rfunction{quantify} method returns \Robject{MSnSet} objects, that 
extend the well-known \Robject{eSet} class defined in the 
\Rpackage{Biobase} package. \Robject{MSnSet} instances are very similar 
to \Robject{ExpressionSet} objects, except for the experiment meta-data 
that captures MIAPE specific information.
The assay data is a matrix of dimensions 
$n \times m$ matrix, where $m$ is the  number of features/spectra 
originally in the \Robject{MSnExp} used as parameter in 
\Robject{quantify} and $m$ is the number of reporter ions, that 
can be accessed with the \Rfunction{exprs} method.
The meta data is directly inherited from the \Robject{MSnExp} instance.

<<quantify,echo=TRUE,cache=FALSE,keep.source=T>>=
qnt <- quantify(experiment,
                method="trap",
                reporters=iTRAQ4,
                verbose=FALSE)
qnt
head(exprs(qnt))
@

>>>>

Unlike the name might suggest, the \Rfunction{removePeaks} method does 
not actually remove peaks from the spectrum; they are set to 0. This 
can be checked using the \Rfunction{peaksCount} method, that returns 
the number of peaks (including 0 intensity peaks) in a spectrum.
To effectively remove 0 intensity 
peaks from spectra, and reduce the size of the data set, one can use 
the \Rfunction{clean} method. 
The effect of the \Rfunction{removePeaks} and \Rfunction{clean} methods 
are illustrated on figure \ref{fig:preproc} on page \pageref{fig:preproc}.

@@

<<preprosp,cache=FALSE,echo=FALSE>>=
int <- c(0,1,1,3,1,1,0,0,0,1,3,7,3,1,0)
mz <- c(113.9,114.0,114.05,114.1,114.15,114.2,114.25,
        114.3,114.35,114.4,114.42,114.48,114.5,114.55,114.6)
ppsp <- new("Spectrum2",intensity=int,mz=mz,centroided=FALSE)
p1 <- plot(ppsp,full=TRUE) + theme_gray(5) + 
  geom_point(size=3,alpha=I(1/3)) +
  geom_hline(yintercept=3,linetype=2) + 
  opts(title="Original spectrum") 
p2 <- plot(removePeaks(ppsp,t=3),full=TRUE) + 
  theme_gray(5) +
  geom_point(size=3,alpha=I(1/3)) + 
  geom_hline(yintercept=3,linetype=2) + 
  opts(title="Peaks < 3 removed") 
p3 <- plot(clean(removePeaks(ppsp,t=3)),full=TRUE) + 
  theme_gray(5) +
  geom_point(size=3,alpha=I(1/3)) + 
  geom_hline(yintercept=3,linetype=2) + 
  opts(title="Peaks < 3 removed and cleaned") 
@

\begin{figure}[p]
  \begin{center}
<<preproc-plot,fig=T,echo=F,tikz=T,external=F,width=3,height=6>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(3, 1)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
print(p1, vp=vplayout(1,1))
print(p2, vp=vplayout(2,1))
print(p3, vp=vplayout(3,1))
@ 
\caption{This figure illustrated the effect or the \Rfunction{removePeaks} 
  and \Rfunction{clean} methods. The left-most spectrum displays two peaks, 
  of max height 3 and 7 respectively. The middle spectrum shows the result 
  of calling \Rfunction{removePeaks} with argument \Robject{t=3}, which sets 
  all data points of the first peak, whose maximum height is smaller or equal 
  to \Robject{t} to 0. The second peak is unaffected. Calling \Rfunction{clean} 
  after \Rfunction{removePeaks} effectly deletes successive 0 intensities from 
  the spectrum, as shown on the right plot. }
\label{fig:preproc}
  \end{center}
\end{figure}

Another useful manipulation method is \Rfunction{trimMz}, that takes as 
parameters and \Robject{MSnExp} (or a \Robject{Spectrum}) and a numeric 
\Robject{mzlim}. MZ values smaller then \texttt{min(mzlim)} 
or greater then \texttt{max(mzmax)} are discarded. This method is 
particularely useful when one wants to concentrate on a specific MZ range, 
as for reporter ions quantification, and generally results in substantial 
reduction of data size. Compare the size of the full trimmed experiment to 
the original \Sexpr{sz} Mb.
\label{trimMz-example}

@@

Quantitation is performed on specific peaks in the spectra, that are specified 
with an \Robject{ReporterIon} object. A specific peak is defined by it's expected 
\Robject{mz} value and is expected to be found within \Robject{mz} $\pm$ \Robject{width}.  
If no data is found, \Robject{NA} is returned.

@@

The \Rfunction{quantify} method takes the following paramters: 
an \Robject{MSnExp} experiment, 
a character describing the quantification \Robject{method}, 
the \Robject{reporters} to be quantified 
and a \Robject{strict} logical defining whether data points 
ranging outside of \Robject{mz} $\pm$ \Robject{width} should be considered 
for quantitation. 
Additionally, a progress bar can be dispaying when setting 
the \Robject{verbose} parameter to \Robject{TRUE}.
Three quantification methods are implemented, as illustrated 
on figure \ref{fig:quant-methods}: 
\Robject{trapezoidation} returns the area under the peak of interest, 
\Robject{max} returns the apex of the peak and 
\Robject{sum} returns the sum of all intensities of the peak. 
See \texttt{?quantify} for more details.

<<simplesp,cache=FALSE,echo=FALSE>>=
int <- c(0,1,1,3,1,1,0)
mz <- c(113.9,114.0,114.05,114.1,114.15,114.2,114.25)
ssp <- new("Spectrum2",intensity=int,mz=mz,centroided=FALSE)
p <- plot(ssp,full=TRUE)
p <- p + theme_gray(5)
@

\begin{figure}[p]
<<quantitation-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=5>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 2)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
print(p + opts(title="Quantitation using 'sum'") + geom_point(size=3,alpha=I(1/3),colour="red"), vp=vplayout(1,1))
print(p + opts(title="Quantitation using 'max'") + geom_point(aes(x=114.1,y=3),alpha=I(1/3),colour="red",size=3), vp=vplayout(1,2))
print(p + opts(title="Trapezoidation and strict=FALSE") + geom_polygon(alpha=I(1/5),fill="red"),vp=vplayout(2,1))
print(p + opts(title="Trapezoidation and strict=TRUE") + 
      geom_polygon(aes(x=c(114.05,114.05,114.1,114.15,114.15),y=c(0,1,3,1,0)),fill="red",alpha=I(1/5)),vp=vplayout(2,2))
@ 
\caption{The different quantitation methods are illustrated above. 
  Quantitation using \Robject{sum} sums all the data points in the peaks to 
  produce, for ths example, \Sexpr{quantify(sp,iTRAQ4[1],method="sum")[[1]]}, 
  whereas method \Robject{max} only uses the peak's maximum intensity, 
  \Sexpr{quantify(sp,iTRAQ4[1],method="max")[[1]]}. 
  \Robject{Trapezoidation} calculates the area under the peak taking 
  the full with into account (using \Robject{strict=FALSE} gives
  \Sexpr{round(quantify(ssp,iTRAQ4[1],method="trap",strict=FALSE)[[1]],3)}) 
  or only the width as defined by the reporter 
  (using \Robject{strict=TRUE} gives
  \Sexpr{round(quantify(ssp,iTRAQ4[1],method="trap",strict=TRUE)[[1]],3)}). }
\label{fig:quant-methods}
\end{figure}

\bigskip

The \Rfunction{quantify} method returns \Robject{MSnSet} objects, that 
extend the well-known \Robject{eSet} class defined in the 
\Rpackage{Biobase} package. \Robject{MSnSet} instances are very similar 
to \Robject{ExpressionSet} objects, except for the experiment meta-data 
that captures MIAPE specific information.
The assay data is a matrix of dimensions 
$n \times m$, where $m$ is the  number of features/spectra 
originally in the \Robject{MSnExp} used as parameter in 
\Robject{quantify} and $m$ is the number of reporter ions, that 
can be accessed with the \Rfunction{exprs} method.
The meta data is directly inherited from the \Robject{MSnExp} instance.

<<quantify,echo=TRUE,cache=FALSE,keep.source=T>>=
qnt <- quantify(experiment,
                method="trap",
                reporters=iTRAQ4,
                strict=FALSE,
                verbose=FALSE)
qnt
head(exprs(qnt))
@
